Composing Robustly Safe Compilers

### 2023-06-08 popl/main.tex
Draft Intro

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[acmsmall,review,screen]{acmart}

\usepackage{listings}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Matthis Kruse}
\authornote{Both authors contributed equally to this research.}
\email{matthis.kruse@cispa.de}
\orcid{0000-0003-4062-9666}
\affiliation{%
  \institution{CISPA Helmholtz Center for Information Security}
  \streetaddress{Stuhlsatzenhaus 5}
  \city{Saarbr{\"u}cken}
  \state{Saarland}
  \country{Germany}
  \postcode{66123}
}

@@

  Lastly, parts of our results are verified in Coq.
\end{abstract}

@@

%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
%\begin{teaserfigure}
%  \includegraphics[width=\textwidth]{sampleteaser}
%  \caption{Seattle Mariners at Spring Training, 2010.}
%  \Description{Enjoying the baseball game from the third-base
%  seats. Ichiro Suzuki preparing to bat.}
%  \label{fig:teaser}
%\end{teaserfigure}

%\received{20 February 2007}
%\received[revised]{12 March 2009}
%\received[accepted]{5 June 2009}

@@

\begin{lstlisting}[language=c,caption=``Wrong bounds check of two 64-bit integers.'']
if at <= bounds {
  x[at] = 42;
}
\end{lstlisting}
Note the subtle bug of using \verb|<=| instead of \verb|<|, leading to an out-of-bounds access whenever \verb|at = bounds|.
Furthermore, when \verb|at| and \verb|bounds| are 64-bit integers on a 32-bit architecture, the comparison may not be performed in constant-time: The compiler may bail out as soon as the lower 32 bits are unequal, not bothering to compare the higher 32 bits.

\section{Compositionality}

\section{Robust MS Preservation}

\section{Robust MS+CCT Preservation}

\section{Related Work}

\section{Conclusion}

%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

>>>>

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[acmsmall,review,screen,dvipsnames]{acmart}

\usepackage[colorinlistoftodos]{todonotes}
\usepackage{cleveref}
\usepackage{glossaries}
\usepackage{listings}
\usepackage{xspace}

\input{cmds.tex}

\loadglsentries{acronyms}
\makeglossaries

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Matthis Kruse}
% \authornote{Both authors contributed equally to this research.}
\email{matthis.kruse@cispa.de}
\orcid{0000-0003-4062-9666}
\affiliation{%
  \institution{CISPA Helmholtz Center for Information Security}
  \streetaddress{Stuhlsatzenhaus 5}
  \city{Saarbr{\"u}cken}
  \state{Saarland}
  \country{Germany}
  \postcode{66123}
}

@@

  Lastly, parts of our results are verified in Coq.

\begin{center}\small\it
	{In the following, we use syntax highlighting accessible to both colourblind and black \& white readers~\citep{patrignani2020use}. Specifically, we use a $\src{\text{blue}}$,
	$\src{sans\text{-}serif}$ font for $\src{L_{TMS}}$,
	$\trg{bold}$ font for $\trg{L}$ and $\trg{L_{MS}}$, and $\obj{green}$ font for $\obj{L_{sCCT}}$.
	}
\end{center}
\end{abstract}

@@

@@

\gls{cct} code does not leak information through timing analyses.
%However, it is possible to write \gls{cct} code in a more high-level language like FaCT~\cite{cauligi2019fact}.
It is folklore that programmers prefer to write code in higher-level languages such as \verb|C|.
But, high-level languages do not run on common hardware, so they need to be translated by a compiler.
Due to their optimizing nature, compilers can break security properties, such as \gls{cct}, when performing this translation.
An example of this exists in literature~\cite{barthe2018sec}:

\begin{minipage}{0.49\textwidth}
  \begin{lstlisting}[language=c,caption={Before Compilation},numbers=left]
// x = (b ? x : y)
int mask = -b;
x = (y & mask) | (x & ~mask);
  \end{lstlisting}
\end{minipage}%
\begin{minipage}{0.49\textwidth}
  \begin{lstlisting}[language=c,caption={After Compilation},numbers=left]
if(b) {
  x = y;
}
  \end{lstlisting}
\end{minipage}

\noindent
Here, the compiler recognizes the pattern in the bit hacks and rewrites them to code branching on \verb|b|, violating \gls{cct}.
Because of this, typically, \gls{cct} code is written in assembly.
Thanks to DSLs like FaCT~\cite{cauligi2019fact}, however, this is not necessary.

But, FaCT's compiler gives no guarantees with regards to robustness.
That is, the code may not be \gls{cct} when linked with untrusted third-party libraries.
%For example, a simple table lookup may be fetched from one of many cache layers.
%The higher up layers lead to slower access times which is detectable from just observing the program's %execution time.
%With that, an attacker can identify the entry in the table that we wanted to access.
%A \gls{cct} table lookup has to go through all elements of that table in the same order.
%
Compilers giving such a guarantee are known to be robustly preserving~\cite{abate2019jour}.
These \emph{secure compilers} give a strong formal guarantee for security, since an arbitrary target-level attacker has no chance of violating our property of interest, such as \gls{cct}.\MK{cite CCT compiler that is robustly preserving}

For code to be \gls{cct}, \gls{ms} is a prerequisite, since information is easily leaked in a memory-unsafe world by just reading out of bounds.
The importance of \gls{ms} has given rise to lots of possible countermeasures.
Particularily interesting for this paper are \emph{compiler-based} countermeasures like CETS~\cite{nagarakatte2010cets}, SoftBound~\cite{nagarakatte2009soft}, or MSWasm~\cite{michael2023mswasm}, which describes a \emph{robust enforcement} of \gls{ms}.
Given \gls{cct} requires \gls{ms}, it is natural to combine compiler-based techniques to ensure both \gls{cct} and \gls{ms} hold.
Unfortunately, the composition of secure compilers is unclear.

This paper develops the theory of secure compiler compositions and introduces a secure compilation chain for a combination of \gls{ms} and \gls{cct}.
We do this via 4 contributions:

\contribution{\contribnum Composition of Properties (\Cref{sec:compprop})}
We present a formal framework to reason about compositions of safety properties.
First, we assume there is a most general trace semantics in which a property can be defined and that there is a monitor for that property.
The monitor is defined on a restricted trace semantics without unneccessary information.
The composition of trace properties is then modelled by the composition of monitors.
The two-level approach, i.e., the two different trace semantics allow for a modular extension without touching any already existing definition.

\contribution{\contribnum Composition of Secure Compilers (\Cref{sec:compcomp})}
We introduce a formal framework for the composition of secure compilers.
Composition of secure compilers allows for modular proof engineering, since our framework can be used even if the compilers robustly preserve different properties.
A simple corollary of our main composition theorem is the fact that the composition of secure compilers with equal source and target programming lanugages can be freely swapped without compromising security.
In turn, formal compiler engineers can optimize the compilation tool chain with regards to the phase ordering problem.

\contribution{\contribnum Compiler Robustly Preserving \gls{ms} and \gls{cct} (\Cref{sec:casestud})}
We introduce a case-study consisting of three secure compilers.
To this end, we present the language $\src{L_{TMS}}$ which has a type system with linear constraints to enforce \gls{tms} statically.
%We prove that any well-typed $\src{L_{TMS}}$ program is temporal memory safe.
Then, we introduce $\trg{L}$, a language without a type system, and a compiler $\llbracket\src{\bullet}\rrbracket^{\src{L_{TMS}}\to\trg{L}}$ from $\src{L_{TMS}}$ to $\trg{L}$ that robustly preserves \gls{tms}.
To get full \gls{ms}, we define another compiler $\llbracket\trg{\bullet}\rrbracket^{\trg{L}\to\trg{L_{MS}}}$ which introduces bounds-checks and compose it with $\llbracket\src{\bullet}\rrbracket^{\src{L_{TMS}}\to\trg{L}}$.
Lastly, we define $\obj{L_{sCCT}}$, a language with access to a \emph{model specific register} to enable \gls{cct} execution, and a compiler $\llbracket\irl{\bullet}\rrbracket^{\trg{L_{MS}}\to\obj{L_{sCCT}}}$ which we compose with the other ones.
Finally, this gives us a three-step compilation chain with both \gls{ms} and \gls{cct}.

\contribution{\contribnum Coq Formalisation}
Parts of our results are formalized in the Coq proof assistant.
We use the symbol \CoqSymbol to demonstrate the fact that a definition or a proof is done in Coq.
Since our semantics and compiler definitions are extractable from Coq to Ocaml, we can run them.

We now describe the other sections in the paper.
\Cref{sec:background} establishes the notions of traces, (hyper-)properties, and robust preservation, while \Cref{sec:relwork} compares this work with others.
We conclude in \Cref{sec:concl} and give an outlook for future work.

\contribution{Open Source \& Technical Report} Our technical report and the Coq formalisation are
available as supplementary material and will be made available upon acceptance of this paper.

\section{Background}\label{sec:background}
\section{Composing Properties}\label{sec:compprop}
\section{Composing Robustly Secure Compilers}\label{sec:compcomp}
\section{Case Study: Memory Safe and Cryptographic Constant Time Code}\label{sec:casestud}

%\begin{lstlisting}[language=c,caption=``Wrong bounds check of two 64-bit integers.'']
%if at <= bounds {
%  x[at] = 42;
%}
%\end{lstlisting}
%Note the subtle bug of using \verb|<=| instead of \verb|<|, leading to an out-of-bounds access whenever \verb|at = bounds|.
%Furthermore, when \verb|at| and \verb|bounds| are 64-bit integers on a 32-bit architecture, the comparison may not be performed in constant-time: The compiler may bail out as soon as the lower 32 bits are unequal, not bothering to compare the higher 32 bits.

\section{Related Work}\label{sec:relwork}
\section{Conclusion}\label{sec:concl}

%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{main}
